---
title: "Non-BRCA-Study-Info"
author: "Nick R. Bachelder"
date: "4/21/2022"
output: html_document
---




```{r setup, include=FALSE}
## Run BRCA study before running this script to get proper global environment

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(readxl)
library(zoo)
library(dplyr)
library(tidyr)
library(openxlsx)
library(ggplot2)
library(survival)
library(glmnet)
library(survminer)
library(kableExtra)
library(stringr)

load('/Users/nickbachelder/Desktop/Duke Internship 2022/brca-negative-stud/all_data2.RData')
```



### Data summary

```{r}
column <- c('Unique Visits for Needle Asperation', 'Unique Patients in Study')
attributes <- c('Number in Original Study', 'Number in Final Sample', 'Removed for missing DOB', 'Removed for missing appointment date',
                'Removed for BRCA positive',
                'Removed for having diagnosis before appointments', 'Removed for last clinic date before RPFNA Date', 'Removed for incorrect appointment date', 
                'Removed for incompleteness of covariates', 'Patients Lost from Original Study')
numbers_app <- c(n_app_before_any_filter, num_final_app, n_app_missing_DOB, n_app_missing_Date, n_app_BRCA, rm_app_after_diag, rm_app_issue_last, app_rm_time, completeness_rm_app,
                 n_app_before_any_filter - num_final_app)
numbers_pat <- c(n_pat_before_any_filter, num_final_pat, n_pat_missing_DOB, n_pat_missing_Date, n_pat_BRCA, rm_pat_after_diag, rm_pat_issue_last,  pat_rm_time, completeness_rm_pat,
                 n_pat_before_any_filter - num_final_pat)
x <- data.frame(numbers_app, numbers_pat) %>% `colnames<-`(column) %>% `rownames<-`(attributes) %>% kbl() %>% kable_classic() 

kableExtra::row_spec(x, c(2, 8), extra_css = "border-bottom: 1px solid")
```

### Missing patient needle asperation data by appointment

```{r}
missing_Date %>% kbl() %>% kable_classic() 
```

### Missing patient date of birth

```{r}
missing_DOB %>% kbl() %>% kable_classic() 
```


### Patients with appointments after diagnosis

```{r, echo=FALSE, warning = FALSE}
after_diag %>% kbl() %>% kable_classic() 
```

### Patients with RPFNA appointments after supposed last clinic visit

```{r, echo=FALSE, warning = FALSE}
issues_last %>% kbl() %>% kable_classic() 
```


### Patients with appointment time issues
```{r}
time_issue %>% kbl() %>% kable_classic() 
```


### Patients that were BRCA positive
```{r}
BRCA_pos %>% kbl() %>% kable_classic() 
```




### Final data sample
```{r}
non_BRCA_modeling_final[1:20,]  %>% kbl() %>% kable_classic() 
```




## Covariance of variables

```{r}
library(corrplot)
df <- non_BRCA_modeling_final %>% dplyr::select(Start_Age, BMI, Highest_Masood, Duration_Btwn_Appt, Atypia_Either) %>% mutate(Atypia_Either = as.numeric(Atypia_Either))
cor.table = cor(df)
corrplot(cor.table, method = 'color', order = 'AOE', tl.col = 'black', type = 'lower')
```

### Model with only atypia

```{r}
res.cox <- coxph(Surv(time = Start_Age, time2 = Stop_Age, Diagnosed)  ~ Atypia_Either +  cluster(ids), data = non_BRCA_modeling_final)
summary(res.cox)


new_df <- with(non_BRCA_modeling_final,
               data.frame(ids = c('1', '1'),
                          Atypia_Either = as.factor(c(0, 1)),
                          Duration_Btwn_Appt = rep(mean(Duration_Btwn_Appt, na.rm = T), 2)
                          )
               )
ggsurvplot(survfit(res.cox, newdata = new_df, data = non_BRCA_modeling_final), conf.int = TRUE, legend.labs=c("Atypia_Either=0", "Atypia_Either=1"),
           ggtheme = theme_minimal())
```

### Model with only Race

0: WHite
1: AA
2: Hispanic
3: Missing / NA


# Distribution of RPFNA by School
```{r}
non_BRCA_modeling_final %>% ggplot(aes(x = Date_RPFH, group = Instit, fill = Instit)) + geom_density(alpha = 0.5) + 
  ggtitle('RPFNA Dates OSE v Duke')
```





















































```{r}
res.cox <- coxph(Surv(time = Start_Age, time2 = Stop_Age, Diagnosed)  ~  Race +  cluster(ids), data = non_BRCA_modeling_final)
summary(res.cox)

new_df <- with(non_BRCA_modeling_final,
               data.frame(ids = c('1', '51'),
                          Race = rep(as.factor(c('White','AfricanAmerican')))
                          )
               )
ggsurvplot(survfit(res.cox, newdata = new_df, data = non_BRCA_modeling_final), conf.int = TRUE, legend.labs=c('Race=White', "Race=AA"),
           ggtheme = theme_minimal())
```


### Model with race and atypia


```{r}
res.cox <- coxph(Surv(time = Start_Age, time2 = Stop_Age, Diagnosed)  ~ Atypia_Either + Race +  cluster(ids), data = non_BRCA_modeling_final)
summary(res.cox)

non_BRCA_modeling_final %>% filter(Instit != 'Duke', Diagnosed == 1)
```


### Model with a few of the other continuous variables. NOTE: Highest_Masood significant when not included with Atypia, which it is correlated with.

```{r}
res.cox <- coxph(Surv(time = Start_Age, time2 = Stop_Age, Diagnosed)  ~  Highest_Masood + cluster(ids), data = non_BRCA_modeling_final)
summary(res.cox)
```


```{r}
res.cox <- coxph(Surv(time = Start_Age, time2 = Stop_Age, Diagnosed)  ~  Atypia_Either + Highest_Masood + BMI, data = non_BRCA_modeling_final)
summary(res.cox)
```











